Mendelian randomization for causal inference in genome-wide association studies

### 2022-03-04 mr-review.Rmd
started omitted.Rmd for removed refs

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via MR [@swerdlow_selecting_2016]. A GWAS
interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005].
In the nearly 17 years since publication of early studies, researchers have reported 
thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012]. Ever larger sample sizes in GWAS have enhanced statistical power to detect associations and refined our understanding of 
human health and disease.

@@

@@

Inverse variance weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
median estimation [@bowden_consistent_2016], Bayesian weighted MR 
[@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix 
[@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID 
[@yuan_likelihood_2021] are all recent methods developments. These methods differ in 
their approaches to two challenges:

1. instrument selection (from among the available SNPs)  
1. modeling and controlling for horizontal pleiotropy

Recent methods developments have enabled execution of two-sample MR studies with only 
GWAS summary statistics [@burgess2015using]. These strategies are especially pertinent now that there is 
an abundance of freely available GWAS summary data [@buniello2019nhgri]. The 
ability to use published data has contributed to a rapid increase in the number of
two-sample MR methods for GWAS summary statistics.

@@

Three assumptions are central to MR. Satisfying them enables inference of 
causal effects in an observational MR study. In practice, careful evaluation of the assumptions is needed. 
Because assumptions 2 and 3, which involve absence of certain types of confounding, can't be verified, 
practitioners must be cautious when planning MR analyses. Sensitivity analysis is highly recommended as a 
way to assess robustness of estimates in the presence of possible assumption violations.

@@

Approaches to MR can be classified into four categories: ratio of coefficients method, two-stage methods, 
likelihood-based methods, and semiparametric methods.

@@

Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics 
is the earliest approach for likelihood-based inference in MR [@davidson1993estimation;@hansen1996finite].

@@

@@

@smith_mendelian_2004 recognized that many genetic variants individually explain a 
small proportion of the variation in a trait, and, thus, sufficient statistical 
power would require sample sizes in the tens of thousands 
[@schatzkin2009mendelian;@burgess_mendelian_2013]. @palmer_using_2012 partially 
resolved this issue by proposing use of multiple SNP instruments in a single MR 
analysis. @pierce_power_2011 reasoned that a single causal estimate derived from 
a collection of causal SNPs would have greater precision than an estimate derived 
from only one SNP [@burgess_mendelian_2013]. At the same time, sharing of GWAS 
summary statistics was encouraged as a best practice [@nature2012asking]. 
@harbord_severity_2013 recognized that one could calculate the Wald ratio estimate of 
the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
a strategy for determining a single causal effect from summary statistics for a 
collection of SNPs. @burgess_mendelian_2013 termed their method "inverse variance 
weighting" (IVW) because each causal SNP's effect is weighted by the inverse of the variance 
of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
causal effects.

@@

Genetic instrumental variable regression uses GWAS summary statistics to construct polygenic scores for 
the outcome of interest [@diprete_genetic_2018]. In constructing polygenic scores from subsets of a GWAS 
cohort, @diprete_genetic_2018 aim to reduce bias due to horizontal pleiotropy.

>>>>

Genome-wide association studies (GWAS) present unique opportunities for 
discovery of causal relationships via MR [@swerdlow_selecting_2016]. A GWAS
interrogates millions of single nucleotide polymorphisms (SNPs) to infer which affect the trait [@the_wellcome_trust_case_control_consortium_genome-wide_2007;@klein_complement_2005].
In the nearly 17 years since publication of early studies, researchers have reported 
thousands of novel SNP-trait associations from GWAS [@klein_complement_2005;@welter_nhgri_2014;@visscher_10_2017;@visscher_five_2012]. Ever larger sample sizes in GWAS have enhanced statistical power to detect associations and have refined our understanding of 
human health and disease.

@@

A number of excellent MR reviews have appeared in the literature [@burgess_review_2017;@richmond_concepts_2022;@sanderson_mendelian_2022]. Our review doesn't 
aim to replace them, but seeks to complement their findings. 
We present recent methods developments for an audience of statisticians and epidemiologists. We discuss in detail methods advances before briefly summarizing 
applications and directions for future research. The value of MR analysis with 
GWAS data will continue to grow in parallel with statistical methods advances.

@@

Recent methods developments have enabled execution of two-sample MR studies with only 
GWAS summary statistics [@burgess2015using]. These strategies are especially pertinent now that there is 
an abundance of freely available GWAS summary data [@buniello2019nhgri]. The 
ability to use published data has contributed to a rapid increase in the number of
two-sample MR methods for GWAS summary statistics (Figure \@ref(fig:article-counts)).

@@

Three assumptions are central to MR. Satisfying them enables inference of 
causal effects in an observational MR study. In practice, careful evaluation of the assumptions is needed. 
Because assumptions 2 and 3, which involve absence of certain types of confounding, can't be verified, 
practitioners must be cautious when planning MR analyses. Sensitivity analysis is highly recommended as a 
way to assess robustness of estimates in the presence of possible assumption violations [@vanderweele_methodological_2014].

@@

Approaches to MR with one SNP instrument can be classified into four categories: ratio of coefficients method, two-stage methods, 
likelihood-based methods, and semiparametric methods [@burgess_review_2017]. We discuss each of these before turning to methods 
that

@@

Likelihood-based MR methods, unlike two-stage methods, provide maximum likelihood 
estimates with their many desirable properties [@fisher1925theory]. Limited information maximum likelihood from econometrics 
is the earliest approach for likelihood-based inference in MR [@davidson1993estimation;@hansen1996finite]. We consider it below before introducing a Bayesian method.

@@

Inverse variance weighted MR [@burgess_mendelian_2013], MR-Egger [@bowden_mendelian_2015], weighted 
median estimation [@bowden_consistent_2016], Bayesian weighted MR 
[@zhao_bayesian_2019], robust adjusted profile score [@zhao_statistical_2020], MRMix 
[@qi_mendelian_2019], CAUSE [@morrison_mendelian_2020], and MRAID 
[@yuan_likelihood_2021] are all recent methods developments. These methods differ in 
their approaches to three considerations:

1. instrument selection (from among the available SNPs)  
1. modeling and controlling for horizontal pleiotropy  
1. statistical inference procedure

@smith_mendelian_2004 recognized that many genetic variants individually explain a 
small proportion of the variation in a trait, and, thus, sufficient statistical 
power would require sample sizes in the tens of thousands 
[@schatzkin2009mendelian;@burgess_mendelian_2013]. @palmer_using_2012 partially 
resolved this issue by proposing use of multiple SNP instruments in a single MR 
analysis. @pierce_power_2011 reasoned that a single causal estimate derived from 
a collection of causal SNPs would have greater precision than an estimate derived 
from only one SNP [@burgess_mendelian_2013]. At the same time, sharing of GWAS 
summary statistics was encouraged as a best practice [@nature2012asking]. 
@harbord_severity_2013 recognized that one could calculate the Wald ratio estimate of 
the causal effect from GWAS summary statistics, and @burgess_mendelian_2013 devised 
a strategy for determining a single causal effect from summary statistics for a 
collection of SNPs.

Below, we discuss a number of MR methods that use multiple 
instruments. The overall trend appears to be in the direction of increasingly sophisticated 
modeling of horizontal pleiotropy (and correlated horizontal pleiotropy) while
attempting to maintain scalability and computational efficiency.

@@

@burgess_mendelian_2013 termed their method "inverse variance 
weighting" (IVW) because each causal SNP's effect is weighted by the inverse of the variance 
of the ratio estimator, and the overall causal effect is the sum of the weighted SNP
causal effects.

@@
